---
title: "Chapter 3: Data Visualization"
output: html_notebook
---

Load the tidyverse package
```{r}
library(tidyverse)
```


## First steps

Do cars with big engines use more fuel than cars with small engines?

```{r}
# mpg dataset
mpg
# plot hwy against displ
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))
```

```{r}
ggplot(data = mpg)
# rows mpg
dim(mpg)
# hwy vs cyl
ggplot(data = mpg) + geom_point(mapping = aes(x = cyl, y = hwy))
# class vs drv
ggplot(data = mpg) + geom_point(mapping = aes(x = drv, y = class))

```

## Aesthetic mappings
```{r}
# hwy vs displ coloured by class
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, colour = class))
```

Mapping class to the alpha and shape aesthetic
```{r}
# alpha
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
# shape
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

Manually change aesthetic
```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy), colour = 'blue')
```

Mapping continous variable to colour
```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, colour = cty))
```

Mapping continous variable to shape
```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, shape = cty))
```

Stroke aesthetic
```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, stroke = cty))
```

Mapping boolean to aesthetic
```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, colour = cty > 15))
```

## Facets

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(~ class, nrow = 2)
```

Facet by 2 variables
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(drv ~ cyl)
```

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ drv)
```


Facet on continous variable
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(~ cty)
```

## Geometric objects

Line geometric object
```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

Mapping drv to linetype aesthetic
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
```

Adding points coloured by drv
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = drv)) +
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv, colour = drv))
```

Using `group` to draw multiple rows of data
```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))
```

Adding multiple geoms 
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

Same as above, but using global mapping
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth()
```

Using local mappings to extend or overwritw global mapping
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(colour = class)) +
  geom_smooth()
```

Use subset of data for `geom_smooth` only
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```

Recreate charts in exercise 3.6.1
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth(se = FALSE)

```

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth(mapping = aes(group = drv), se = FALSE)
```

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, colour = drv)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(colour = drv)) +
  geom_smooth(se = FALSE)
```

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(colour = drv)) +
  geom_smooth(mapping = aes(linetype = drv), se = FALSE)
```

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(fill = drv), shape = 21, colour = "white", size = 4, stroke = 2)
```

