---
title: "Chapter 5: Data transformation"
output: html_notebook
---

## Prerequisites

```{r}
library(nycflights13)
library(tidyverse)
```

Base R functions have been overwritten. To access them, use full names: `stats::filter()`, `stats::lag()`

## nycflights13

```{r}
flights
```

## Filter row with filter()

All flights on Jan 1st
```{r}
filter(flights, month == 1, day == 1)
```

Save result in variable

```{r}
(dec25 <-  filter(flights, month == 12, day == 25))
```

## Comparisons

== / =

```{r}
filter(flights, month = 1)
```

Beware floating point numbers
```{r}
sqrt(2) ^ 2 == 2
```

```{r}
1 / 49 * 49 == 1
```

Use near instead

```{r}
near(sqrt(2) ^ 2, 2)
```

```{r}
near(1 / 49 * 49, 1)
```

## Logical Operators

Multiple arguements to filter are combined with `and`.

Find all flights departed in November or December

```{r}
filter(flights, month == 11 | month == 12)
```

Using `%in%` for same results

```{r}
(nov_dec <- filter(flights, month %in% c(11, 12)))
```

Using De Morgan's law to simplify complicated subsetting: `!(x & y)` same as `!x | !y` and `!(x | y)` same as `!x & !y`

```{r}
filter(flights, !(arr_delay > 120 | dep_delay > 120))
filter(flights, arr_delay <= 120, dep_delay <= 120)
```

## Missing values

```{r}
NA > 10
NA == 2
NA * 4
NA == NA
```
Check if value is missing
```{r}
x <- NA
is.na(x)
```

`filter()` only includes rows evaluated with `TRUE`, excludes both `FALSE` and `NA`

```{r}
df <- tibble(x = c(1, NA, 3, 4))
filter(df, x > 2)
```

Explicitly include NA

```{r}
filter(df, is.na(x) | x > 2)
```

## 5.2.4 Exercises

Flights that had an arrival delay of two or more hours

```{r}
filter(flights, arr_delay >= 2)
```

Flights that flew to Houston
```{r}
filter(flights, dest %in% c("IAH", "HOU"))
```

Flights that were operated by United, American or Delta

```{r}
filter(flights, carrier %in% c("UA", "AA", "DL"))
```

