---
title: "Chapter 5: Data Transformation"
author: "Arnold Taremwa"
date: "31 May 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Prerequisites

```{r}
library(nycflights13)
library(tidyverse)
```

Base R functions have been overwritten. To access them, use full names: `stats::filter()`, `stats::lag()`

## nycflights13

```{r}
flights
```

## Filter row with filter()

All flights on Jan 1st
```{r}
filter(flights, month == 1, day == 1)
```

Save result in variable

```{r}
(dec25 <-  filter(flights, month == 12, day == 25))
```

## Comparisons

== / =

```{r}
# error
#filter(flights, month = 1)
```

Beware floating point numbers
```{r}
sqrt(2) ^ 2 == 2
```

```{r}
1 / 49 * 49 == 1
```

Use near instead

```{r}
near(sqrt(2) ^ 2, 2)
```

```{r}
near(1 / 49 * 49, 1)
```

## Logical Operators

Multiple arguements to filter are combined with `and`.

Find all flights departed in November or December

```{r}
filter(flights, month == 11 | month == 12)
```

Using `%in%` for same results

```{r}
(nov_dec <- filter(flights, month %in% c(11, 12)))
```

Using De Morgan's law to simplify complicated subsetting: `!(x & y)` same as `!x | !y` and `!(x | y)` same as `!x & !y`

```{r}
filter(flights, !(arr_delay > 120 | dep_delay > 120))
filter(flights, arr_delay <= 120, dep_delay <= 120)
```

## Missing values

```{r}
NA > 10
NA == 2
NA * 4
NA == NA
```
Check if value is missing
```{r}
x <- NA
is.na(x)
```

`filter()` only includes rows evaluated with `TRUE`, excludes both `FALSE` and `NA`

```{r}
df <- tibble(x = c(1, NA, 3, 4))
filter(df, x > 2)
```

Explicitly include NA

```{r}
filter(df, is.na(x) | x > 2)
```

## 5.2.4 Exercises

Flights that had an arrival delay of two or more hours

```{r}
filter(flights, arr_delay >= 2)
```

Flights that flew to Houston
```{r}
filter(flights, dest %in% c("IAH", "HOU"))
```

Flights that were operated by United, American or Delta

```{r}
filter(flights, carrier %in% c("UA", "AA", "DL"))
```

Departed in summer (July, August, September)

```{r}
filter(flights, month %in% c(7, 8, 9))
```

Arrived more than 2 hours late but didnt leave late

```{r}
filter(flights, arr_delay > 2, dep_delay <= 0)
```

Delayed by at least one hour, but made up over 30 minutes in flight
```{r}

```

Departed between midnight and 6am

```{r}
filter(flights, dep_time >= 0, dep_time <= 600)
```

Use `dplyr`'s `between` to replicate above results

Departed between midnight and 6am
```{r}
filter(flights, between(dep_time, 0, 600))
```

How many flights with missing `dep_time`

```{r}
filter(flights, is.na(dep_time))
```

Other variables with NA

```{r}
colSums(is.na(flights))
```

## Arange rows with `arrange()`

```{r}
arrange(flights, year, month, day)
```

Descending order

```{r}
arrange(flights, desc(arr_delay))
```

NA is always sorted at the end

```{r}
df <- tibble(x = c(3, 2, NA))
arrange(df, x)
arrange(df, desc(x))
```

## 5.3.1 Exercises

Sort NA to the top
```{r}
arrange(df, desc(is.na(x)))
```

Most delayed flights

```{r}
arrange(flights, desc(dep_delay))
```

Earliest departures

```{r}
arrange(flights, dep_delay)
```

Fastest flights

```{r}
arrange(flights, air_time)
```

Longest flights

```{r}
arrange(flights, desc(distance))
```

Shortest flights

```{r}
arrange(flights, distance)
```

## 5.4 Select columns with `select`

Select by name

```{r}
select(flights, year, month, day)
```

Select by range (inclusive)
```{r}
select(flights, year:day)
```

Select all columns except these

```{r}
select(flights, -(year:day))
```

Starts with

```{r}
select(flights, starts_with("dep"))
```

Ends with

```{r}
select(flights, ends_with("time"))
```

Contains

```{r}
select(flights, contains("arr"))
```

Matches regex

```{r}
select(flights, matches("[a-z]+_[a-z]+"))
```

Number range `num_range("x", 1:3)` matches x1, x2, x3

Rename columns

```{r}
rename(flights, this_year = year)
```

Everything helper

```{r}
select(flights, time_hour, air_time, everything())
```

## 5.4.1 Exercises

Select dep_time, dep_delay, arr_time, and arr_delay from flights.
```{r}
select(flights, starts_with("dep"), starts_with("arr"))
```

```{r}
select(flights, dep_time, dep_delay, arr_time, arr_delay)
```

```{r}
select(flights, matches("dep_[a-z]+"), matches("arr_[a-z]+"))
```

Including variable name multiple times

```{r}
select(flights, dep_time, dep_time)
```

`one_of` helper function

```{r}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")

select(flights, one_of(vars))
```

`contains` helper

```{r}
select(flights, contains("TIME"))
```

## Add new variables with mutate()

```{r}
flights_sml <- select(flights, 
  year:day, 
  ends_with("delay"), 
  distance, 
  air_time
)
mutate(flights_sml,
  gain = arr_delay - dep_delay,
  speed = distance / air_time * 60
)
```

Refer to newly created columns

```{r}
mutate(flights_sml,
       gain = arr_delay - dep_delay,
       hours = distance / air_time * 60,
       gain_per_hour = gain / hours
       )
```

To keep new columns, use `transmute`

```{r}
transmute(flights, 
          gain = arr_delay - dep_delay,
          hours = distance / air_time * 60,
          gain_per_hour = gain / hours
          )
```

## Useful creation functions

Modular arithmetic

```{r}
transmute(flights,
  dep_time,
  hour = dep_time %/% 100,
  minute = dep_time %% 100
)
```

Logs: `log()`, `log2()`, `log10()`

Offsets: lead() and lag()

```{r}
(x <- 1:10)
lag(x)
lead(x)
```

Cumulative and rolling packages

```{r}
x
cumsum(x)
cummean(x)
```

Logical Comparisons

Ranking

```{r}
y <- c(1, 2, 3, NA, 3, 4)

min_rank(y)

min_rank(desc(y))

row_number(y)

dense_rank(y)

percent_rank(y)

cume_dist(y)
```

## 5.5.2 Exercises

Convert `dep_time` and `sched_dep_time` to more convinient representation of number of minutes since midnight

```{r}
transmute(flights, 
          dep_time_minutes = ((dep_time %/% 100) * 60) + (dep_time %% 100),
          sched_dep_time_minutes = ((sched_dep_time %/% 100) * 60) + (sched_dep_time %% 100))
```

Compare air_time with arr_time - dep_time

```{r}
transmute(flights,
          air_time,
          calc_air_time = arr_time - dep_time)
```


10 most delayed flights using ranking

```{r}
arrange(flights, min_rank(desc(dep_delay)))
```