---
title: 'Chapter 7: Exploratory Data Analysis'
author: "Arnold Taremwa"
date: "1 June 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exploratory Data Analysis

## Prerequisites

```{r}
library(tidyverse)
library(ggstance)
library(lvplot)
library(ggbeeswarm)
library(hexbin)
```

## 7.3.1 Visualizing distribution

Categorical variable 

```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))
```

```{r}
diamonds %>%
  count(cut)
```

Continuous variable

```{r}
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat), binwidth = 0.5)
```

```{r}
diamonds %>%
  filter(carat < 3) %>%
  ggplot() + 
    geom_histogram(mapping = aes(x = carat), binwidth = 0.1)
```

Grouping

```{r}
diamonds %>%
  filter(carat < 3) %>%
  ggplot() +
    geom_freqpoly(mapping = aes(x = carat, colour = cut))
```

Unusual values

```{r}
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = y), binwidth = 0.5)
```

Zoom in 

```{r}
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = y), binwidth = 0.5) +
  coord_cartesian(ylim = c(0, 60))
```

```{r}
diamonds %>%
  filter(y < 3 | y > 20) %>%
  select(x, y, z) %>%
  arrange(y)
```

## 7.3.4 Exercises 

1. Explore the distribution of each of the x, y, and z variables in diamonds. What do you learn? Think about a diamond and how you might decide which dimension is the length, width, and depth.

```{r}
ggplot(diamonds) +
  geom_histogram(aes(x), binwidth = 0.5) +
  labs(title = "x distribution")
```

```{r}
ggplot(diamonds) +
  geom_histogram(aes(y), binwidth = 0.5) +
  labs(title = "y distribution")
```

```{r}
ggplot(diamonds) +
  geom_histogram(aes(z), binwidth = 0.5) +
  labs(title = "z distribution")
```

2. Explore the distribution of price. Do you discover anything unusual or surprising? (Hint: Carefully think about the binwidth and make sure you try a wide range of values.)

```{r}
ggplot(diamonds) +
  geom_histogram(aes(price), binwidth = 500)
```

```{r}
ggplot(diamonds) +
  geom_histogram(aes(price), binwidth = 50)
```

```{r}
diamonds %>%
  count(cut_width(price, 50)) %>%
  filter(n <= 10)
```

3. How many diamonds are 0.99 carat? How many are 1 carat? What do you think is the cause of the difference?

```{r}
diamonds %>%
  summarise(
    n_0_99 = sum(carat == 0.99),
    n_1 = sum(carat == 1)
  )
```

4. Compare and contrast coord_cartesian() vs xlim() or ylim() when zooming in on a histogram. What happens if you leave binwidth unset? What happens if you try and zoom so only half a bar shows?

```{r}
ggplot(diamonds) +
  geom_histogram(aes(price)) +
  coord_cartesian(xlim = c(0, 5000))

ggplot(diamonds) +
  geom_histogram(aes(price)) +
  xlim(0, 5000)
```


## Missing values

## Covariation

Categorical and continuous

```{r}
ggplot(data = diamonds, mapping = aes(x = price)) + 
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 500)

# density plot | area under curve = 0
ggplot(data = diamonds, mapping = aes(x = price, y = ..density..)) + 
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 500)
```

The boxplot

```{r}
ggplot(diamonds) +
  geom_boxplot(aes(cut, price))
```

Reorder the categories

```{r}
ggplot(mpg) +
  geom_boxplot(aes(class, hwy))
```

```{r}
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy))

```

## 7.5.1.1 Exercises

1. Use what you’ve learned to improve the visualisation of the departure times of cancelled vs. non-cancelled flights.


```{r}
nycflights13::flights %>%
  mutate(cancelled = if_else(is.na(dep_delay), "cancelled", "not cancelled")) %>%
  mutate(sched_dep_time_hr = hour + minute / 60) %>%
  ggplot() +
    geom_boxplot(mapping = aes(x = cancelled, y = sched_dep_time_hr))

```

2. What variable in the diamonds dataset is most important for predicting the price of a diamond? How is that variable correlated with cut? Why does the combination of those two relationships lead to lower quality diamonds being more expensive?

```{r}

```

3. Install the ggstance package, and create a horizontal boxplot. How does this compare to using coord_flip()?

```{r}
ggplot(mpg) +
  geom_boxplot(mapping = aes(x = class, y = hwy)) +
  coord_flip()

```

```{r}
ggplot(mpg) +
  geom_boxploth(mapping = aes(x = hwy, y = class))
```

4. One problem with boxplots is that they were developed in an era of much smaller datasets and tend to display a prohibitively large number of “outlying values”. One approach to remedy this problem is the letter value plot. Install the lvplot package, and try using geom_lv() to display the distribution of price vs cut. What do you learn? How do you interpret the plots?

```{r}
ggplot(data = diamonds) +
  geom_boxplot(mapping = aes(x = cut, y = price))

ggplot(data = diamonds) +
  geom_lv(mapping = aes(x = cut, y = price))
```

5. Compare and contrast geom_violin() with a facetted geom_histogram(), or a coloured geom_freqpoly(). What are the pros and cons of each method?

Violin

```{r}
ggplot(data = diamonds) +
  geom_violin(mapping = aes(x = cut, y = price)) 
```

Facetted histogram

```{r}
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = price)) +
  facet_wrap(~ cut, nrow = 1)
```

Colured frequency

```{r}
ggplot(data = diamonds) +
  geom_freqpoly(mapping = aes(x = price, colour = cut))
```

6. If you have a small dataset, it’s sometimes useful to use `geom_jitter()` to see the relationship between a continuous and categorical variable. The `ggbeeswarm` package provides a number of methods similar to `geom_jitter()`. List them and briefly describe what each one does.

```{r}
ggplot(data = mpg) +
  geom_jitter(mapping = aes(x = cty, y = hwy))
```

```{r}
ggplot(data = mpg) +
  geom_beeswarm(mapping = aes(x = cty,y = hwy))
```

```{r}
ggplot(data = mpg) +
  geom_quasirandom(mapping = aes(x = cty,y = hwy))
```

Two categorical variables

Counts for each combination
```{r}
ggplot(data = diamonds) +
  geom_count(mapping = aes(x = cut, y = color))
```

```{r}
diamonds %>%
  count(color, cut)
```

```{r}
diamonds %>%
  count(color, cut) %>%
  ggplot(mapping = aes(x = color, y = cut)) +
    geom_tile(mapping = aes(fill = n))
```

## 7.5.2.1 Exercises

1. How could you rescale the count dataset above to more clearly show the distribution of cut within colour, or colour within cut?

```{r}
diamonds %>%
  count(color, cut)
  
```

2. Use geom_tile() together with dplyr to explore how average flight delays vary by destination and month of year. What makes the plot difficult to read? How could you improve it?

```{r}
nycflights13::flights %>%
  filter(!is.na(dep_delay)) %>%
  select(month, dest, dep_delay) %>%
  group_by(month, dest) %>%
  summarise(
    avg_delay = mean(dep_delay)
  ) %>%
  ggplot(mapping =aes(x = month, y = dest)) +
    geom_tile(mapping = aes(fill = avg_delay))
```

3. Why is it slightly better to use aes(x = color, y = cut) rather than aes(x = cut, y = color) in the example above?



Two continuous variables

Scatterplot

```{r}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price))
```

```{r}
ggplot(data = diamonds) + 
  geom_point(mapping = aes(x = carat, y = price), alpha = 1 / 100)
```

```{r}
smaller <- diamonds %>%
  filter(carat < 3)
ggplot(data = smaller) +
  geom_bin2d(mapping = aes(x = carat, y = price))

ggplot(data = smaller) +
  geom_hex(mapping = aes(x = carat, y = price))
```

Group a continous variable

```{r}
ggplot(data = smaller, mapping = aes(x = carat, y = price)) + 
  geom_boxplot(mapping = aes(group = cut_width(carat, 0.1)))
```

Make boxplot width proportional to sqrt of number of observations

```{r}
ggplot(data = smaller, mapping = aes(x = carat, y = price)) + 
  geom_boxplot(mapping = aes(group = cut_width(carat, 0.1)), varwidth = TRUE)
```

Put equal number of observations in each bin

```{r}
ggplot(data = smaller, mapping = aes(x = carat, y = price)) + 
  geom_boxplot(mapping = aes(group = cut_number(carat, 20)))
```

## 7.5.3.1 Exercises

1. Instead of summarising the conditional distribution with a boxplot, you could use a frequency polygon. What do you need to consider when using cut_width() vs cut_number()? How does that impact a visualisation of the 2d distribution of carat and price?

```{r}
ggplot(data = smaller, mapping = aes(x = price)) + 
  geom_freqpoly(mapping = aes(y = ..density.., colour = cut_width(carat, 0.1)))
```

```{r}
ggplot(data = smaller, mapping = aes(x = price)) + 
  geom_freqpoly(mapping = aes(y = ..density.., colour = cut_number(carat, 20)))
```

2. Visualise the distribution of carat, partitioned by price.

```{r}
ggplot(data = smaller, mapping = aes(x = price, y = carat)) + 
  geom_boxplot(mapping = aes(group = cut_number(price, 20)))
```

3. How does the price distribution of very large diamonds compare to small diamonds. Is it as you expect, or does it surprise you?

```{r}
diamonds %>%
  mutate(volume = x * y * z) %>%
  filter(volume < 1000) %>%
  ggplot(mapping = aes(x = volume, y = price)) +
    geom_boxplot(mapping = aes(group = cut_width(volume, 50))) +
    geom_smooth(se = FALSE)
```

```{r}
ggplot(data = smaller, mapping = aes(x = carat, y = price)) +
  geom_point(alpha = 1/10) +
  geom_smooth(se = FALSE)
```


4. Combine two of the techniques you’ve learned to visualise the combined distribution of cut, carat, and price.

```{r}
ggplot(data = smaller) +
  geom_bar(mapping = aes(x = cut_width(carat, 1), y = price, fill = cut),
           stat = "identity")
```

5. 


